---
title: "Data Analysis of the gapminder dataset"
author: "Vittorio Romaniello"
date: "05/09/2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
options(width = 10)
knitr::opts_chunk$set(echo = TRUE, comment=NA, size = "small")
```

## Introduction
Here we present the results of the analysis of the gapminder dataset.

The gapminder dataset contains data on life expectancy, GDP per capita and population by country.

More information can be found at <http://www.gapminder.org/data/>.

## General Analysis 
```{r echo=FALSE}
library(gapminder)
library(knitr)
```
The dataset contains ``r nrow(gapminder)`` rows and ``r ncol(gapminder)`` columns.

In line with the description of the dataset, the columns are:

```{r, echo=FALSE}
names(gapminder)
```

The number of unique countries in the dataset is ``r length(unique(gapminder$country))``.

Observations are collected over the years: 
```{r echo=FALSE, results='hold'}
options(width = 90)
unique(gapminder$year)
```

## Analysis by variable
Continuing our analysis, we will focus on the variables:

* `Population`
* `Life Expectancy`
* `GDP per capita`

## Population

China is the country with the largest population, followed by India.
```{r echo=FALSE}
options(width = 70)
kable(gapminder[gapminder$pop == max(gapminder$pop), c("country", "continent", "year", "pop")])


kable(gapminder[gapminder$country != "China" & gapminder$pop == max(gapminder[gapminder$country != "China", c("pop")]), c("country", "continent", "year", "pop")])
```

## Population

The country with the smallest population is Sao Tome and Principe.
```{r echo=FALSE}
kable(gapminder[gapminder$pop == min(gapminder$pop), c("country", "continent", "year", "pop")])
```



Sao Tome and Principe is also the country with the smallest population in 2007.
```{r, echo=FALSE}
kable(gapminder[gapminder$pop == min(gapminder[gapminder$year == 2007, c("pop")]), c("country", "continent", "year", "pop")])
```

## Population

Increasing trend over the years.
```{r echo=FALSE}
boxplot(log(pop)~year,data=gapminder, xlab="Year", ylab = "log(Population)", main="Boxplot of log(Population) by year")
```

## Life expectancy

* Highest: Japan (2007)
* Lowest Rwanda (1992)

```{r, echo=FALSE}
kable(gapminder[gapminder$lifeExp == max(gapminder$lifeExp),])

kable(gapminder[gapminder$lifeExp == min(gapminder$lifeExp),])

```

## Life expectancy 

Significant change in life expectancy. Median life span increased by approximately 25 years, from 45 years in 1952 to 70 years in 2007.

```{r echo=FALSE}
boxplot(lifeExp~year, data=gapminder, xlab="Year", ylab="Life Expectancy", main="Boxplot of life expectancy by year")
```

## GDP per capita

- Richest country Kuwait (1957), GDP per capita:``r as.integer(gapminder[gapminder$gdpPercap == max(gapminder$gdpPercap), c("gdpPercap")])`` US$. 
- Poorest country Congo (2002), GDP per capita ``r as.integer(gapminder[gapminder$gdpPercap == min(gapminder$gdpPercap), c("gdpPercap")])`` US$.

GDP per capita has increased over the years.
```{r echo=FALSE}
boxplot(log(gdpPercap)~year, data=gapminder, xlab="Year", ylab="Log(GDP)", main="Boxplot of log(GDP) by year")
```

## Conclusion

### Gapminder Dataset

* Life expectancy, GDP per capita and population have increased over the years.
* Countries with highest/lowest GDP per capita are not also the countries with highest/lowest life expectancy.
  * Life expectancy is determined also by other factors than a country's GDP.

